Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2019-02-25 pkg/disttree/inst/draft/draft.Rnw
removed highlighting of changes and unneeded comments; final formatting
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2474 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0be09dda9fd42d8a35ac2a25b130c9625d684d82

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Subsampling is employed 
  for each tree using a third of the predictors in each split of the tree
  (\hlc{argument} \code{mtry} \hlc{in our implementation} \code{distforest},
  \hlc{with more ``computational details'' provided at the end of the manuscript}).
  Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}. 
  The stopping criteria are the minimal number of observations to perform a split 
  (\texttt{minsplit} $= 50$), the minimal number of observations in a 
  segment (\texttt{minbucket} $= 20$), and the significance level 
  for variable selection (\texttt{alpha} $= 1$). The latter means
  that no early stopping (or ``pre-pruning'') is applied
  based on results of the statistical tests.
  % stopping criteria:
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

@@

In a first step, we show a detailed comparison of the competing models
for one observation site, Axams in Tyrol (in the center of the study area,
see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation 
observations and numerical weather predictions are available for the month of July 
from 1985 through 2012. In Figure~\ref{fig:axams-dist} in the introduction the 
probabilistic forecasts from the distributional forest, trained on 1985--2008, 
for July 24 in 2009--2012 have already been shown as a motivational example. 
\hlc{In particular, the figure depicts the forecasted point mass at zero
(i.e., the probability of a dry day) along with the forecasted probability 
density function for the total amount of precipitation. Based on this illustration 
it can be observed that} the four sample forecasts differ considerably in 
location $\mu$, scale $\sigma$, and the amount of censoring while conforming quite 
well with the actual observations from these days. 
While this is a nice illustrative example we are  
interested in the overall predictive performance and calibration of the
distributional fits. More details of this assessment as well as
an application to 14~further meteorological stations is provided in
Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b}.

@@

To assess whether these differences in predictive performance are due to differences
in the topography, Figure~\ref{fig:all-map} shows a brief spatial summary of all
stations. Each station is illustrated by a symbol that conveys which
model performed best in terms of CRPS on the last 4~years of the data.
Additionally, the color of the symbol indicates the CRPS difference between 
distributional forest and the best-performing other model. Green signals that
the distributional forest performs better than the other models whereas red
signals that another model performs better. Overall the distributional forest
performs on par (gray) or better (green) for the majority of stations. Only
for a few stations in the north-east EMOS performs best, and in East Tyrol
the prespecified GAMLSS performs particularly well in the validation period
(2009--2012). \hlc{Partially, this can be attributed to random variation as
the differences at several stations are mitigated when considering a full
cross-validation rather than a single split into learning and validation period}
\citep[see Supplement~B,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019b}.
\hlc{Further differences are possibly due to East Tyrol lying in a different
climate zone, south of the main Alpine Ridge. Hence, long-term climatological
characteristics as well as the precipitation patterns in 2009--2012 differ
from North Tyrol, conforming particularly well with the additive effects from
the prespecified GAMLSS.}

@@

\hlc{Beyond the task of modeling precipitation it is of interest how well distributional
forests perform in combination with other more general distributional specifications.
It has been shown previously in the literature that using a score- or gradient-based
selection of splitting variables outperforms a mean-based selection with subsequent
flexible distributional modeling: For example, both} \citet[][Figure~2]{Athey+Tibshirani+Wager:2019}
\hlc{and} \citet[][Figure~1]{Hothorn+Zeileis:2017} \hlc{demonstrated (independently) that their
respective score-based random forest algorithms outperform the mean-based quantile
regression forests of} \cite{Meinshausen:2006} \hlc{in a setup where only the variance of a normal
response variable changes across the considered covariates. However, if all distribution
parameters are closely correlated with the distribution mean the forests with different
splitting strategies all perform similarly, provided a sufficiently flexible distribution
is employed for the final predictions} \citep[see][Section~7]{Hothorn+Zeileis:2017}.

\hlc{Similarly, the score-based distributional forests introduced in this manuscript
proved to be quite robust to the different distributional specifications considered.
While all specifications focus on capturing mean-variance effects note that these
parameters are never fully orthogonal but can actually become quite closely correlated
due to the censoring (or truncation and/or zero-inflation considered in} 
Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a}).

\hlc{However, exploring extensions to more flexible parametric distributions}
\citep[e.g., such as the Dagum distribution considered by][in GAMLSS-type models]{Klein+Kneib+Lang:2015}
\hlc{as well as transformation model specifications} \citep[e.g., as in][]{Hothorn+Zeileis:2017}
\hlc{are of interest for future research.}

@@

Selecting tuning parameters for flexible regression models is important
not only in terms of predictive accuracy but also computational complexity. 
For the application in Section~\ref{sec:precipitation} tuning parameters
are selected based on advice from the literature as well as our own experiences.
As \cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} recommend to 
build full-grown trees, early stopping upon non-significance is disabled
(\texttt{alpha} $= 1$) and low values are used for \texttt{minsplit} ($ = 50$)
and \texttt{minbucket} ($ = 20$), while assuring that \texttt{minsplit}
is sufficiently large for reasonably obtaining MLEs of all parameters in each
segment of the tree.

@@

\section*{Acknowledgements}
\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
2017 to January 2018) was financially supported by the Swiss National
Science Foundation, grant number SNF IZSEZ0 177091.}

\newpage
\section*{Appendix}
\begin{appendix}

@@

\cite{Strasser+Weber:1999} \hlc{showed that the asymptotic conditional 
distribution of the linear statistic} $t_l$ \hlc{is a multivariate normal with parameters}
$\mu$ and $\Sigma$. \hlc{Hence, the asymptotic conditional distribution of}
$c(t_{l},\mu_{l},\Sigma_{l})$ 
\hlc{is either normal}
(for $c_{\text{max}}$) 
\hlc{or} 
$\chi^2$ (for $c_{\text{quad}}$).

@@

@@

\newpage
\bibliography{ref.bib}

>>>>

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Subsampling is employed 
  for each tree using a third of the predictors in each split of the tree
  (argument \code{mtry} in our implementation \code{distforest},
  with more ``computational details'' provided at the end of the manuscript).
  Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}. 
  The stopping criteria are the minimal number of observations to perform a split 
  (\texttt{minsplit}~$= 50$), the minimal number of observations in a 
  segment (\texttt{minbucket}~$= 20$), and the significance level 
  for variable selection (\texttt{alpha}~$= 1$). The latter means
  that no early stopping (or ``pre-pruning'') is applied
  based on results of the statistical tests.
  % stopping criteria:
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

@@

In a first step, we show a detailed comparison of the competing models
for one observation site, Axams in Tyrol (in the center of the study area,
see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation 
observations and numerical weather predictions are available for the month of July 
from 1985 through 2012. In Figure~\ref{fig:axams-dist} in the introduction the 
probabilistic forecasts from the distributional forest, trained on 1985--2008, 
for July 24 in 2009--2012 have already been shown as a motivational example. 
In particular, the figure depicts the forecasted point mass at zero
(i.e., the probability of a dry day) along with the forecasted probability 
density function for the total amount of precipitation. Based on this illustration 
it can be observed that the four sample forecasts differ considerably in 
location $\mu$, scale $\sigma$, and the amount of censoring while conforming quite 
well with the actual observations from these days. 
While this is a nice illustrative example we are  
interested in the overall predictive performance and calibration of the
distributional fits. More details of this assessment as well as
an application to 14~further meteorological stations is provided in
Supplement~B \citep{Schlosser+Hothorn+Stauffer:2019b}.

@@

To assess whether these differences in predictive performance are due to differences
in the topography, Figure~\ref{fig:all-map} shows a brief spatial summary of all
stations. Each station is illustrated by a symbol that conveys which
model performed best in terms of CRPS on the last 4~years of the data.
Additionally, the color of the symbol indicates the CRPS difference between 
distributional forest and the best-performing other model. Green signals that
the distributional forest performs better than the other models whereas red
signals that another model performs better. Overall the distributional forest
performs on par (gray) or better (green) for the majority of stations. Only
for a few stations in the north-east EMOS performs best, and in East Tyrol
the prespecified GAMLSS performs particularly well in the validation period
(2009--2012). Partially, this can be attributed to random variation as
the differences at several stations are mitigated when considering a full
cross-validation rather than a single split into learning and validation period
\citep[see Supplement~B,][and the corresponding discussion in the next section]{Schlosser+Hothorn+Stauffer:2019b}.
Further differences are possibly due to East Tyrol lying in a different
climate zone, south of the main Alpine Ridge. Hence, long-term climatological
characteristics as well as the precipitation patterns in 2009--2012 differ
from North Tyrol, conforming particularly well with the additive effects from
the prespecified GAMLSS.

@@

Beyond the task of modeling precipitation it is of interest how well distributional
forests perform in combination with other more general distributional specifications.
It has been shown previously in the literature that using a score- or gradient-based
selection of splitting variables outperforms a mean-based selection with subsequent
flexible distributional modeling: For example, both \citet[][Figure~2]{Athey+Tibshirani+Wager:2019}
and \citet[][Figure~1]{Hothorn+Zeileis:2017} demonstrated (independently) that their
respective score-based random forest algorithms outperform the mean-based quantile
regression forests of \cite{Meinshausen:2006} in a setup where only the variance of a normal
response variable changes across the considered covariates. However, if all distribution
parameters are closely correlated with the distribution mean the forests with different
splitting strategies all perform similarly, provided a sufficiently flexible distribution
is employed for the final predictions \citep[see][Section~7]{Hothorn+Zeileis:2017}.

Similarly, the score-based distributional forests introduced in this manuscript
proved to be quite robust to the different distributional specifications considered.
While all specifications focus on capturing mean-variance effects note that these
parameters are never fully orthogonal but can actually become quite closely correlated
due to the censoring (or truncation and/or zero-inflation considered in
Supplement~A \citep{Schlosser+Hothorn+Stauffer:2019a}).

However, exploring extensions to more flexible parametric distributions
\citep[e.g., such as the Dagum distribution considered by][in GAMLSS-type models]{Klein+Kneib+Lang:2015}
as well as transformation model specifications \citep[e.g., as in][]{Hothorn+Zeileis:2017}
are of interest for future research.

@@

Selecting tuning parameters for flexible regression models is important
not only in terms of predictive accuracy but also computational complexity. 
For the application in Section~\ref{sec:precipitation} tuning parameters
are selected based on advice from the literature as well as our own experiences.
As \cite{Hastie+Tibshirani+Friedman:2001} and \cite{Breiman:2001} recommend to 
build full-grown trees, early stopping upon non-significance is disabled
(\texttt{alpha}~$= 1$) and low values are used for \texttt{minsplit}~($ = 50$)
and \texttt{minbucket}~($ = 20$), while assuring that \texttt{minsplit}
is sufficiently large for reasonably obtaining MLEs of all parameters in each
segment of the tree.

@@

%\newpage
\section*{Appendix}
\begin{appendix}

@@

\cite{Strasser+Weber:1999} showed that the asymptotic conditional 
distribution of the linear statistic $t_l$ is a multivariate normal with parameters
$\mu$ and $\Sigma$. Hence, the asymptotic conditional distribution of
$c(t_{l},\mu_{l},\Sigma_{l})$ 
is either normal
(for $c_{\text{max}}$) 
or 
$\chi^2$ (for $c_{\text{quad}}$).

@@

\section*{Acknowledgments}
An extended research stay of Torsten Hothorn in Innsbruck (August
2017 to January 2018) was financially supported by the Swiss National
Science Foundation, grant number SNF IZSEZ0 177091.

@@

%\newpage
\bibliography{ref.bib}
